A Test of the Frequency Independence Assumption of Power System Coefficients used in Geomagnetically Induced Current Estimates

### 2019-07-10 paper.tex
Updates from Overleaf

\begin{abstract}
A common assumption used when estimating geomagnetically induced currents (GICs) in a power system given a time series of nearby direct measurements or indirect estimates of the horizontal geoelectric field components $E_x(t)$ and $E_y(t)$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency-independent power system coefficients.  A first test of this assumption uses GIC measurements made in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$9~km away. A second model (Model 2) is obtained using a frequency domain generalization of Model 1: $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$. The coefficients $a(\omega)$ and $b(\omega)$ are computed from data and are shown to be frequency dependent and this model provides significantly better estimates of the measured GIC than Model~1. Based on results using a simulated geoelectric field, it is suggested that the measurement-derived frequency dependence of the system coefficients may be explained by spatial variations in the spectrum of the geoelectric field along the line in which GIC is measured rather than by an actual frequency dependence of the system parameters. It is also shown that further improvements over Model~2 can be made using a model with the geomagnetic field as an input.
\end{abstract}

@@

In this work, we use a unique dataset in which time series measurements of $\mathbf{E}$ and the surface horizontal geomagnetic field, $\mathbf{B}$, were made near a site where GIC was being measured in an electric power system.  We first estimate the system coefficients $a_o$ and $b_o$ from data using conventional methods, in which they are modeled as frequency independent. Then, we use a model in which the system coefficients are allowed to be frequency dependent, compute the frequency dependence, and compare this model's ability to represent GIC with the traditional frequency-independent model.

@@

The 1-second-cadence GIC data used span a total of 35 days across 10 intervals that start and end at midnight Japan Standard Time (JST) and are listed in Table~\ref{intervals}. These data have been presented previously in \citet{Watari2009} and \cite{Watari2015}. These time intervals are those for which GIC data was made available with the exception of three days for which there were obvious problems with either the electric field or GIC measurements. The GIC dataset includes 1~s raw data and 1~Hz low-pass-filtered raw measurements. The results presented here are for the 1~Hz low-pass-filtered GIC measurements

@@

Data for the time span of 2006-08-04T15:00 through 2006-08-08T15:00 are shown Figure~\ref{sample}. The GIC data were provided as 1-day files that start at midnight JST and the times shown in Figure~\ref{sample} are Universal Time.

Spikes in the electric field measurements were assumed to be unphysical when the absolute value of a component $i$ of the electric field changed by more than 1~mV/km at time $t$, and the values $E_i(t-2)$ through $E_i(t+4)$ were replaced using linear interpolation of the values at $E_i(t-3)$ and $E_i(t + 5)$. In the sample shown in Figure~\ref{sample}, 21 spikes were identified in $E_x$ and zero in $E_y$. For all of the time intervals considered, the fraction of interpolated electric field values ranged from 0--0.1\%.

@@

The GIC data contains non-physical spikes that are followed by 40-80~s of anomalous values that appear as an exponential decay back to pre-spike values. These spikes were removed by identifying times $t$ when the change in the 1~s absolute value of $GIC(t)$ was greater than $0.05$~Amps and then values in a window $GIC(t-2)$ through $GIC(t+100)$ were replaced using a linear interpolation of the values at $GIC(t-3)$ and $GIC(t + 101)$. In the example interval shown in Figure~\ref{sample}, 493 spikes were removed. For all of the time intervals considered, the fraction of interpolated GIC values ranged from 1--4\%.

@@

In Model~2, the system coefficients $a(\omega)$ and $b(\omega)$ are frequency dependent according to

\begin{linenomath*}
\begin{equation}
G_E(\omega) = a(\omega)E_x(\omega) + b(\omega)E_y(\omega)
\label{model2}
\end{equation}
\end{linenomath*}
\noindent
and are estimated using the Fourier transforms of  $GIC(t)$ and $\mathbf{E}(t)$; Frequency-dependent variables are indicated by explicitly listing their dependence on $\omega$. The method of parameter estimation and the generation of time-domain GIC estimates for Models~2-4 are given at the end of this section.

@@

\noindent
In this model, the transfer function components $z_x$ and $z_y$ are determined directly from measurements of $GIC(t)$ and $\mathbf{B}(t)$ and no electric field measurements are used; this model is expected to produce estimates of GIC that are similar to that obtained from Model~3 because they both use the same input variable.

Estimates of $a(\omega)$ and $b(\omega)$ are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing a ordinary least squares regression (OLS) using Equation~\ref{model2} and a set $G_E(\omega)$ and $\mathbf{E}(\omega)$ in a band of frequencies around $f_e$ from the DFT (discrete Fourier transform) of 24 hours of 1-s cadence data. Following the suggestion of \cite{Simpson2005}, the 28 values of $f_e$ start at $0.25$ Hz and decrease by a factor of approximately $\sqrt{2}$ and end at $1/43200$~Hz; if a $f_e$ value computed in this way is not an integer multiple of $1/86400$, the nearest smaller integer multiple is used. The periods associated with the evaluation frequencies are listed in Table~\ref{evaluationperiods}. The regression for each $f_e$ uses all DFT frequencies with values in a window with a width of twice the separation from the next lower value of $f_e$ and the given $f_e$ with the exceptions that the lowest evaluation frequency band has a lower bound of $1/86400$~Hz and highest evaluation frequency band has an upper bound of $0.375$~Hz.

@@

Transfer function coefficients were computed for each of the segments. The average transfer function magnitude and phase at a given evaluation frequency was obtained by averaging the segment values at that evaluation frequency.

To interpolate the frequency-domain transfer function coefficients estimated at the evaluation frequencies, their real and imaginary components are linearly interpolated individually onto the original DFT frequency grid (with a spacing of $1/86400$~Hz), and then the inverse DFT is computed. The interpolated transfer function coefficients were set to zero at frequencies higher than $0.25$~Hz and the transfer function estimate at $f=0$ was set to zero before interpolation.

@@

Several modifications to this basic procedure were considered. Robust regression \citep{Egbert1986} with a Huber weighting constant of $1.345$ and an upper limit of 50 iterations and bi-square weighting with a constant of $4.685$ (both constants provide 95\% efficiency for normally distributed errors), both with no hard-cutoff step; time domain pre-whitening and tapering; and weighting the frequencies around each evaluation frequency using Parzen window weights. When using these additional steps individually, we find that the conclusions made regarding the differences each model's prediction ability do not change because the differences in each model's prediction performance are within the calculated uncertainty of the used method. Transfer function coefficients found using these modifications are generally within the error bars of those for the used procedure when the signal to noise ratio is above $\sim$3 (corresponding to periods longer than $\sim30$~s; At periods below $\sim$30~s, the electric field measurements are significantly influenced by the instrument response characteristics~\citep{Fujii2015}, aliasing, and the computed values are also expected to be significantly biased due to low signal to noise ratio.)

Out-of-sample estimation metrics for each model on day $D$ were determined by computing the average model coefficients using the remaining 34 days of data and then using these averaged coefficients to generate GIC time series and associated metrics on day $D$. This process was repeated to create a set of 35 out-of-sample prediction metrics for each model. In calculating the metrics, the first and last 10 minutes of the day were excluded to reduce the influence of transients that exist in the estimate at the start of the interval for the amount of time that the causal part of the impulse response (found by Fourier inversion of the transfer function) is non-zero and at the end for the amount of time that the acausal part of the impulse response is non-zero.

The performance of each model in estimating GIC was assessed using three metrics: the prediction efficiency, PE, the mean-squared-error (MSE) ratio, and the signal-to-noise ratio. The PE and MSE ratios are shown in Table~\ref{results}.

@@

@@

Figure~\ref{predictions} shows the predictions for each model in a 1-day interval selected because the value of the prediction efficiency for each model was in the same order as that shown in Table~\ref{results}, i.e., Model~1 has the lowest PE and Model~4 has the highest. Note that the large errors visible at the start and end of the time intervals are due to the transients discussed above and the first and last 10~minutes of each interval were omitted from metrics calculations.

@@

To determine the statistical significance of the differences, $10^5$ bootstrap averages of the segment PEs of each model were created and their differences were computed. From this, the null hypothesis that the PE of Model~2 is equal to that of Model~1 can be rejected at a significance level of less than $10^{-5}$ because no bootstrap average PE difference from Model~2 was found to be lower than that for Model~1. The null hypothesis that the PE of Model~3 is equal to that of Model~2 can also be rejected at the same significance level. The null hypothesis that the PE of Model~4 is equal to that of Model~3 can be rejected at a significance level of $10^{-2}$.

@@

Figure~\ref{Z_MT} and~\ref{Phi_MT} show the parameters for $\mathbf{Z}$ in Model~3. Note that electric and magnetic field data from Memambetsu, Japan were also used by \cite{Fujii2015} to compute $\mathbf{Z}$. When using the \cite{Fujii2015} transfer function for Model~3, the PE was significantly lower (0.58 vs. 0.78). (The \cite{Fujii2015} transfer function was obtained from the IRIS DMC EMTF database [\cite{Kelbert2011}]). Although not critical to the issues addressed in this work, a preliminary investigation suggests that the cause is due to several factors. We have found that when using continuous one-month intervals prior to 2006 to compute $\mathbf{Z}$, our calculated transfer function matches the \cite{Fujii2015} transfer function to within the error bars except below periods of 100~s, which is likely explained by the issues with the electric field response in the 1-second-cadence measurements discussed in \cite{Fujii2015}. After 2006, our calculated transfer functions differ significantly from that in \cite{Fujii2015} - they are shifted downward by approximately 10-50\%. In addition, we have found that the estimated transfer function is sensitive to whether or not segment averaging was used. Due to the large gaps between and the varying length of continuous segments of data used in this work, transfer function coefficients were computed using one-day segments of data and then averaged. We have found that if a continuous set of measurements over approximately 21 days is used to compute a single transfer function, the transfer function magnitudes are up to 25\% larger than if 21 transfer functions were computed and then averaged. This result is unexpected, and extensive successful verification tests of our MT calculations along with the fact that this effect has observed by the co-author, who used different MT software on different data sets, suggests the explanation is not due to an error in the code used her to compute transfer functions. We expect to consider this in a future investigation as it suggests that stacking can be a significant source of transfer function bias and note that the most likely explanation involves non-stationarity in the noise.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/plot_raw_All_20060805.pdf}
\caption{(Top) Geomagnetic field measured at the Memanbetsu Magnetic Observatory (MMB). (Middle) Geoelectric field measured at the Memanbetsu Magnetic Observatory (MMB), which is $\sim$9 km from the Memanbetsu substation. (Bottom) GIC measurements measured in a grounded neutral point of a Y-connected three-phase transformer connected to a 187 kV bus at the Memanbetsu substation of the Hokkaido Power Co. The subscripts $x$ and $y$ correspond to Geographic North and East, respectively.}
\label{sample}
\end{figure}

>>>>

\begin{abstract}
A common assumption used when estimating geomagnetically induced currents (GICs) in a power system given a time series of nearby direct measurements or indirect estimates of the horizontal geoelectric field components $E_x(t)$ and $E_y(t)$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency-independent power system coefficients.  A first test of this assumption is made using GIC measurements in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$9~km away. A second model (Model 2) is obtained using a frequency domain generalization of Model 1: $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$. The coefficients $a(\omega)$ and $b(\omega)$ are computed from data and are shown to be frequency dependent and this model provides significantly better estimates of the measured GIC than Model~1. Based on results using a simulated geoelectric field, it is suggested that the measurement-derived frequency dependence of the system coefficients may be explained by spatial variations in the spectrum of the geoelectric field along the line in which GIC is measured rather than by an actual frequency dependence of the system components. It is also shown that further improvements over Model~2 can be made using models with the geomagnetic field as an input.
\end{abstract}

@@

In this work, we use a unique dataset in which time series measurements of $\mathbf{E}$ and the surface horizontal geomagnetic field, $\mathbf{B}$, were made near a site where GIC was being measured in an electric power system.  We first estimate the system coefficients $a_o$ and $b_o$ from data using conventional methods, in which they are modeled as frequency independent. Then, we use a model in which the system coefficients are allowed to be frequency dependent, compute the frequency dependence, and compare this model's ability to represent GIC with the traditional frequency-independent model. Two additional models are considered that require only the geomagnetic field as an input and they are shown to be significantly better at estimating GIC than the models that use only the geoelectric field as an input.

@@

The 1-second-cadence GIC data used span a total of 35 days across 10 intervals that start and end at midnight Japan Standard Time (JST) and are listed in Table~\ref{intervals}. These data along with a map showing the location of the transformer and the connected 187~kV power line have been presented previously in \citet{Watari2009} and \cite{Watari2015}. These time intervals are those for which GIC data was made available with the exception of three days for which there were obvious problems with either the electric field or GIC measurements. The GIC dataset includes 1~s raw data and 1~Hz low-pass-filtered raw measurements. The results presented here are for the 1~Hz low-pass-filtered GIC measurements.

@@

Data for the time span of 2006-08-04T15:00Z through 2006-08-08T15:00Z are shown Figure~\ref{sample}. The GIC data were provided as 1-day files that start at midnight JST and the times shown in Figure~\ref{sample} are Universal Time.

Spikes in the electric field measurements were assumed to be unphysical when the absolute value of a component $i$ of the electric field changed by more than 1~mV/km at time $t$, and the values $E_i(t-2)$ through $E_i(t+4)$ were replaced using linear interpolation of the values at $E_i(t-3)$ and $E_i(t + 5)$. In the sample shown in Figure~\ref{sample}, 21 spikes were identified in $E_x$ and zero in $E_y$. For the time intervals considered, the fraction of interpolated electric field values ranged from 0--0.1\%.

@@

The GIC data contains non-physical spikes that are followed by 40-80~s of anomalous values that appear as an exponential decay back to pre-spike values. These spikes were removed by identifying times $t$ when the change in the 1~s absolute value of $GIC(t)$ was greater than $0.05$~Amps and then values in a window $GIC(t-2)$ through $GIC(t+100)$ were replaced using a linear interpolation of the values at $GIC(t-3)$ and $GIC(t + 101)$. In the example interval shown in Figure~\ref{sample}, 493 spikes were removed. For the time intervals considered, the fraction of interpolated GIC values ranged from 1--4\%.

@@

In Model~2, the system coefficients $a(\omega)$ and $b(\omega)$ are frequency dependent according to (frequency-dependent variables are indicated by explicitly listing their dependence on $\omega$)

\begin{linenomath*}
\begin{equation}
G_E(\omega) = a(\omega)E_x(\omega) + b(\omega)E_y(\omega)
\label{model2}
\end{equation}
\end{linenomath*}
\noindent
and were estimated using the Fourier transforms of  $GIC(t)$ and $\mathbf{E}(t)$; the method of coefficient estimation and the generation of time-domain GIC estimates for this and the following models are given at the end of this section.

@@

\noindent
In this model, the transfer function components $z_x$ and $z_y$ are determined directly from measurements of $GIC(t)$ and $\mathbf{B}(t)$ and no electric field measurements are used; this model is expected to produce estimates of GIC that are similar to that obtained from Model~3 because they both use the same input measurements.

Estimates of $a(\omega)$ and $b(\omega)$ are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing a ordinary least squares regression (OLS) using Equation~\ref{model2} and a set $G_E(\omega)$ and $\mathbf{E}(\omega)$ in a band of frequencies around $f_e$ from the DFT (discrete Fourier transform) of 24 hours of 1-s cadence data. Following the suggestion of \cite{Simpson2005}, the 28 values of $f_e$ start at $0.25$ Hz and decrease by a factor of $\sqrt{2}$ and end at $1/43200$~Hz; if a $f_e$ value computed in this way is not an integer multiple of $1/86400$, the nearest smaller integer multiple is used. The periods associated with the evaluation frequencies are listed in Table~\ref{evaluationperiods}. The regression for each $f_e$ uses all DFT frequencies with values in a window with a width of twice the separation from the next lower value of $f_e$ and the given $f_e$ with the exceptions that the lowest evaluation frequency band has a lower bound of $1/86400$~Hz and highest evaluation frequency band has an upper bound of $0.375$~Hz.

@@

Transfer functions were computed for each of the 1-day segments. The average transfer function magnitude and phase at a given evaluation frequency was obtained by averaging the segment values at that evaluation frequency.

To interpolate the frequency-domain transfer function components estimated at the evaluation frequencies, their real and imaginary components were linearly interpolated individually onto the original DFT frequency grid (with a spacing of $1/86400$~Hz), and then the inverse DFT was computed. The interpolated transfer function coefficients were set to zero at frequencies higher than $0.375$~Hz and the transfer function estimate at $f=0$ was set to zero before interpolation.

@@

Several modifications to this basic procedure were considered. Robust regression \citep{Egbert1986} with a Huber weighting constant of $1.345$ and an upper limit of 50 iterations and bi-square weighting with a constant of $4.685$ (both constants provide 95\% efficiency for gaussian-distributed errors), both with no hard-cutoff step; time domain pre-whitening and tapering; and weighting the frequencies around each evaluation frequency using Parzen window weights. When using these additional steps individually, we find that the conclusions made regarding the differences each model's prediction ability do not change because the differences in each model's prediction performance are within the calculated uncertainty of the used method. Transfer function coefficients found using these modifications are generally within the error bars of those for the used procedure when the signal to noise ratio is above $\sim$3 (corresponding to periods longer than $\sim$30~s; At periods below $\sim$30~s, the electric field measurements are significantly influenced by the instrument response characteristics~\citep{Fujii2015}, aliasing, and the computed values are also expected to be significantly biased due to a low signal-to-noise ratio.)

\section{Model Evaluation}

Out-of-sample estimation quality metrics for each model on day $D$ were determined by computing the average model coefficients using the remaining 34 days of data and then using these averaged coefficients to generate GIC time series and associated metrics on day $D$. This process was repeated to create a set of 35 out-of-sample prediction metrics for each model. In calculating the metrics, the first and last 10 minutes of the day were excluded to reduce the influence of transients that exist in the estimate at the start of the interval for the amount of time that the causal part of the impulse response (found by Fourier inversion of the transfer function) is non-zero and at the end for the amount of time that the acausal part of the impulse response is non-zero.

The performance of each model in estimating GIC was assessed using three metrics: the prediction efficiency, PE, the mean-squared-error (MSE) ratio, and the signal-to-noise ratio (SN).

@@

The PE and MSE ratios are shown in Table~\ref{results} and Figure~\ref{predictions} shows the predictions for each model in a 1-day interval selected because the value of the prediction efficiency for each model was in the same order as that shown in Table~\ref{results}, i.e., Model~1 has the lowest PE and Model~4 has the highest. Note that the large errors visible at the start and end of the time intervals are due to the transients discussed above and the first and last 10~minutes of each interval were omitted from metrics calculations.

@@

@@

To determine the statistical significance of the PE differences, $10^5$ bootstrap averages of the segment PEs of each model were created and their differences were computed. From this, the null hypothesis that the PE of Model~2 is equal to that of Model~1 can be rejected at a significance level of less than $10^{-5}$ because no bootstrap average PE difference from Model~2 was found to be lower than that for Model~1. The null hypothesis that the PE of Model~3 is equal to that of Model~2 can also be rejected at the same significance level. The null hypothesis that the PE of Model~4 is equal to that of Model~3 can be rejected at a significance level of $10^{-2}$.

@@

Figure~\ref{Z_MT} and~\ref{Phi_MT} show the parameters for $\mathbf{Z}$ in Model~3. Note that electric and magnetic field data from Memambetsu, Japan were also used by \cite{Fujii2015} to compute $\mathbf{Z}$. When using the \cite{Fujii2015} transfer function for Model~3, the PE was significantly lower (0.58 vs. 0.78). (The \cite{Fujii2015} transfer function was obtained from the IRIS DMC EMTF database [\cite{Kelbert2011}]). Although not critical to the conclusions in this work, a preliminary investigation suggests several explanations. We have found that when using continuous one-month intervals prior to 2006 to compute $\mathbf{Z}$, our calculated transfer function matches the \cite{Fujii2015} transfer function to within the error bars except below periods of 100~s, which is likely explained by the issues with the electric field response in the 1-second-cadence measurements discussed in \cite{Fujii2015}. After 2006, our calculated transfer functions differ significantly from that in \cite{Fujii2015} - they are shifted downward by approximately 10-50\%. In addition, we have found that the estimated transfer function is sensitive to whether or not segment averaging was used. Due to the large gaps between and the varying length of continuous segments of data used in this work, transfer function parameters were computed using one-day segments of data and then averaged. We have found that if a continuous set of measurements over approximately 21 days is used to compute a single transfer function, the transfer function magnitudes are up to 25\% larger than if 21 transfer functions were computed and then averaged. This result is unexpected, and extensive successful verification tests of our MT calculations along with the fact that this effect has observed by both authors using two different MT calculation codes on different data sets (the other code was MT software distributed with MT hardware), suggests the explanation is not due to an error in the code used her to compute transfer functions. We expect to consider this in a future investigation as it suggests that stack averaging can be a significant source of transfer function bias and note that the most likely explanation involves non-stationarity in the noise.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/plot_raw_All_20060805.pdf}
\caption{(Top) Geomagnetic field measured at the Memanbetsu Magnetic Observatory (MMB). (Middle) Geoelectric field measured at the Memanbetsu Magnetic Observatory (MMB), which is $\sim$9 km from the Memanbetsu substation. (Bottom) GIC measurements measured in a grounded neutral point of a Y-connected three-phase transformer connected to a 187 kV bus at the Memanbetsu substation of the Hokkaido Power Co. The subscripts $x$ and $y$ correspond to Geographic North and East, respectively, and the date labels are for Universal Time.}
\label{sample}
\end{figure}
